---
title: "cvcrand package"
author: "Hengshi Yu, John A. Gallis, Fan Li, and  Elizabeth L. Turner"
date: "`r Sys.Date()`"
bibliography: references.bib
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{cvcrand package for the design and analysis of cluster randomized trials}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---
```{r setup,echo=FALSE,results="hide"}

library(cvcrand)


```


## Overview
cvcrand is an R package for the design and analysis of cluster randomized trials (CRTs). 

Given the baseline values of some cluster-level covariates, users can perform a constrained randomization on the clusters into two arms, with an optional input of user-defined weights on the covariates. 

At the end of the study, the individual outcome is collected. The `cvcrand` package also performs clustered permutation test on either continuous outcome or binary outcome adjusted for some individual-level covariates, producing p-value of the intervention effect.




## Design: constrained randomization

In the design of CRTs with two arms, users can use the `cvcrand()` function to perform constrained randomization. And for the analysis part, user will use the `cptest()` function for clustered permutation test. 

A cluster is the unit of randomization for a cluster randomized trial. Thus, when the number of clusters is small, there might be some baseline imbalance from the randomization between the arms. Constrained randomization constrained the randomization space to randomization schemes with smaller difference among the covariates between the two arms. 

The balance score for constrained randomization in the program is developed from [@raab2001balance]. Suppose $n$, $n_T$, and $n_C$ are the total number of clusters, the number of clusters in the treatment arm and the control arm respectively. Suppose also that there are $K$ cluster-level variables including the continuous covariates as well as the dummy variables created from the categorical covariates. $\bar{x}_{Tk}$ and $\bar{x}_{Ck}$ are the means of the $kth$ variable in the treatment arm and the control arm respectively. $\omega_k$ is the inverse of the standard deviation of the $kth$ variable among all $n$ clusters. There are two choices of metric for the balance score. If `"L1"` metric is specified and there is no user-defined `weight` of $C_k;\ k=1,2,...,K$, the balance score is defined as follows. 

$B_{(l1)}=\left(\frac{n_Tn_C}{n}\right)\sum_{k=1}^{K}\omega_k\left|\bar{x}_{Tk}-\bar{x}_{Ck}\right|$

And if there is user-defined `weight` of $C_k;\ k=1,2,...,K$, the `"L1"` balance score is:
$B_{(l1)}=\left(\frac{n_Tn_C}{n}\right)\sum_{k=1}^{K}C_k\omega_k\left|\bar{x}_{Tk}-\bar{x}_{Ck}\right|$. 

Another metric is `"L2"`. The balance score with and without the user-defined `weight` is defined as follows:

$B_{(l2)}=\left(\frac{n_Tn_C}{n}\right)^{2}\sum_{k=1}^{K}\omega_k^2(\bar{x}_{Tk}-\bar{x}_{Ck})^2$ and $B_{(l2)}=\left(\frac{n_Tn_C}{n}\right)^{2}\sum_{k=1}^{K}C_k^2\omega_k^2(\bar{x}_{Tk}-\bar{x}_{Ck})^2$

With the baseline values of the specified cluster-level covariates in a cluster randomized trail, the `cvcrand()` function in the `cvcrand` package is used to perform the consrailed randomization. 

Each categorical variable is transformed into dummy variables to calculate the balance score. Specifically, the level of each categorical variable dropped when it is transformed into several dummiy variables is the first level in the alphanumerical order. If the user wants to specify a different level of each categorical variable to drop, they might create dummy variables themselves before running the `cvcrand()` function. Then, the user needs to specify the dummy variables created themselves to be `"categorical"` when running `cvcrand()`. 

## `cvcrand()` example

Study 1 in [@dickinson2015pragmatic] is about two approaches (interventions) for increasing the "up-to-date" immunization rate in 19- to 35-month-old children. They planned to randomize 16 counties in Colorado 1:1 to either a population-based approach or a practice-based approach. There are several county-level variables. The program will randomize on a subset of these variables. The continuous variable of average income is categorized to illustrate the use of the `cvcrand()` on multi-category variables. And the percentage in CIIS variable is trancated at 100%.


```{r, echo=FALSE, results='asis'}
knitr::kable(Dickinson_design)
```


For the constrained randomization, we used the `cvcrand()` function to randomize 8 out of the 16 counties into the practice-based. For the definition of the whole randomization space, if the total number of all possible schemes is smaller than `50,000`, we enumerate all the schemes as the whole randomization space. Otherwise, we simulate `50,000` schemes and choose the unique shemes among them as the whole randomization space. We calculate the balance scores of `"L2"` metric on three continuous covariates as well as two categorical covariates of location and income category. Location has `"Rural"` and `"Urban"`. The level of `"Rural"` was then dropped in `cvcrand()`. As income category has three levels of `"low"`, `"med"`, and `"high"`,  the level of `"high"` was dropped to create dummy variables according to the alphanumerical order as well. Then we constrained the randomization space to the schemes with `"L2"` balance scores less than the `0.1` quantile of that in the whole randomization space. Finally, a randomization scheme is sampled from the constrained space. 

We saved the constrained randomization space in a CSV file in `"dickinson_constrained.csv"`, the first column of which is an indicator variable of the finally selected scheme (`1`) or not (`0`). We also saved the balance scores of the whole randomization space in a CSV file in `"dickinson_bscores.csv"`, and output a histogram displaying the distribution of all balance scores with a red line indicating our selected cutoff (the `0.1` quantile).  



```{r cvcrand, fig.keep="all", fig.width = 7, fig.height=4}

 Design_result<-cvcrand(id = Dickinson_design$county,
                       metric = "L2",
                       x = data.frame(Dickinson_design[, -1]),
                       n = 16,
                       ntrt = 8,
                       categorical = c("location","incomecat"),
                       savedata = "dickinson_constrained.csv",
                       savebscores = "dickinson_bscores.csv",
                       cutoff = 0.1,
                       seed = 12345)
 
 


```


The we had the following output: 
```{r set-options, echo=FALSE, fig.keep="all", fig.width = 7, fig.height=4}
options(width = 100)
 
```

```{r , fig.keep="all", fig.width = 7, fig.height=4}
 # the metric the user specified
 Design_result$metric

 # the selected scheme 
 Design_result$Allocation
 
 # the cutoff balance score, the balance score of the selected   scheme, as well as the histogram of the balance scores of the whole randomization space
 Design_result$Bscores
 
 # the statement about how many clusters to randomize to the intervention and the control arms respectively
 Design_result$assignment_message
 
 # the statement about how to get the whole space of schemes
 Design_result$scheme_message
 
 # the statement about the benchmark of the constrained space
 Design_result$BLcut_message
 
 # the statement about the selected scheme from constrained randomization
 Design_result$BLchoice_message
 
 
 # the matrix containing allocation scheme, the id as well as the original covariates' matrix
 Design_result$data_BL
 
 # the descriptive statistics for all the variables in the original covariates' matrix in the two arms from constrained randomization
 Design_result$BL_result


```

From the output of `Design_result$BL_result`, the selected scheme is able to properly balance the baseline values of the covariates. And the selected scheme is shown in `Design_result$Allocation`. 




## Analysis: Clustered Permutation Test

At the end of cluster randomized trials, individual outcomes are collected. Permutation test based on [@gail1996design] and [@li2016evaluation] is then applied to the continuous or binary outcome with some individual-level covariates. 

The `cptest()` function in the `cvcrand` package is used to perform the permutation test for the intervention effect of cluster randomized trials. 

Each categorical variable is transformed into dummy variables to fit in the linear model or logistic regression for the permutation test. Specifically, the level of each categorical variable dropped when it is transformed into several dummiy variables is the first level in the alphanumerical order. If the user wants to specify a different level of each categorical variable to drop, they might create dummy variables themselves before running the `cptest()` function. Then, the user needs to specify the dummy variables created themselves to be `"categorical"` when running `cptest()`. 

## `cptest()` example

Suppose that the researchers were able to assess 300 children in each cluster of the study 1 in [@dickinson2015pragmatic], and the cluster randomized trial is processed with the selected randomization scheme from the example above of the `cvcrand()` function. We expanded the values of the cluster-level covariates on the covariates' values of the individuals, according to which cluster they belong to. The correlated individual outcome of up-to-date on immunizations (`1`) or not (`0`) is then simulated using a generalized linear mixed model (GLMM) to induce correlation by include a random effect. The intracluster correlation (ICC) was set to be 0.01, using the latent response definition provided in [@eldridge2009intra]. This is a reasonable value of the ICC the population health studies [@hannan1994parameters]. We simulated one data set, with the outcome data dependent on the county-level covariates used in the constrained randomization design and a positive treatment effect so that the practice-based intervention increases up-to-date immunization rates more than the community-based intervention. For each individual child, the outcome is equal to `1` if he or she is up-to-date on immunizations and `0` otherwise.

```{r, echo=FALSE, results='asis'}
knitr::kable(head(Dickinson_outcome, 10))
```

We used the `cptest()` function to process the clustered permutation test on the binary outcome of the status of up-to-date on immunizations. We input the file about the constrained space with the first column indicating the final scheme. The permutation test is on the continuous covariates of `"inciis"`, `"uptodateonimmunizations"`, `"hispanic"`, as well as categorical variables of `"location"` and `"incomecat"`. Location has `"Rural"` and `"Urban"`. The level of `"Rural"` was then dropped in `cptest()`. As income category has three levels of `"low"`, `"med"`, and `"high"`,  the level of `"high"` was dropped to create dummy variables according to the alphanumerical order as well.

```{r cptest, fig.keep="all", fig.width = 7, fig.height=4}
 Analysis_result <- cptest(outcome = Dickinson_outcome$outcome,
                           id = Dickinson_outcome$county,
                           x = data.frame(Dickinson_outcome[, c(-1, -7)]),
                           cspacedatname = system.file("dickinson_constrained.csv", package="cvcrand"),
                           outcometype = "binary",
                           categorical = c("location","incomecat"))

``` 

The result of `"cptest()"` includes the final scheme for the cluster randomized trial, the p-value from the permutation test as well as a statement about that p-value.

```{r cptestre, fig.keep="all", fig.width = 7, fig.height=4}
 Analysis_result 
``` 

From the p-value of `0.0497` in `Analysis_result`, the probability of up-to-date on immunizations for the practice-based approach (`1`) is significantly different from that for the population-based approach (`0`). 

# Session Information

```{r info, results='markup', echo=FALSE}
sessionInfo()
```

# References
